\textbf{Evaluation of machine learning methods for 16S rRNA gene data}

### 2019-03-13 submission/manuscript.Rmd
update

@@

# ------- 1. In a loop read the files as delim
# --------2. get the model name for each with get_model_name() 
#---------3. Add a column with model name to each delim
result <- list()
for(file in walltime_files){
  model_walltime <- read_files(file) %>%  
  #summarise_walltime() %>% 
  mutate(model=get_model_name(file))
  result[[length(result)+1]] <- model_walltime
}

# ------- 1. If models are L1 or L2 Linear SVM or L2 Logit then the walltime is in minutes
# --------2. If models are Random Forest or XGBoost then the walltime is in days
#---------3. If other models, the walltime is in hours.
#---------4. This loop converts all to hours.
min_fixed_result <- list()
for(i in result){
  if((sum(i$model=="L1_Linear_SVM")==100 || sum(i$model=="L2_Linear_SVM")==100 || sum(i$model=="L2_Logistic_Regression")==100)){
    i$x <- i$x/60 # The walltimes were saved as minutes # We convert these to hours by diving with 60
  }
  if((sum(i$model=="Random_Forest")==100 || sum(i$model=="XGBoost")==100)){
    i$x <- i$x*24 # The walltimes were saved as minutes # We convert these to hours by diving with 60
  }
  else{
  }
  min_fixed_result[[length(min_fixed_result)+1]] <- i
}

# Bind all delims with model name and walltime together
# Summarise mean walltime
walltime_df <- bind_rows(min_fixed_result) %>% 
  group_by(model) %>% 
  summarise(mean_walltime=mean(x), sd_walltime=sd(x))

@@

Advances in sequencing technology and decreasing costs of generating 16S rRNA gene sequences have allowed rapid exploration of human associated microbiome and its health implications. Currently, the human microbiome field is growing at an unprecedented rate and as a result, there is an increasing demand for methods that identify associations between members of the microbiome and human health. However, this is an undertaking as human associated microbial communities are remarkably complex and uneven. It is unlikely that a single species can explain a disease. Instead, subsets of those communities, in relation to one another and to their host, account for the differences in the health outcomes. Therefore, researchers have started to explore the utility of machine learning (ML) models that use microbiota associated biomarkers to predict human health and to understand the microbial ecology of diseases such as liver cirrhosis, colorectal cancer, inflammatory bowel diseases (IBD), obesity, and type 2 diabetes [@zeller_potential_2014; @zackular_human_2014; @baxter_dna_2016; @baxter_microbiota-based_2016; @hale_shifts_2017; @pasolli_machine_2016; @sze_looking_2016; @walters_meta-analyses_2014; @vazquez-baeza_guiding_2018; @qin_alterations_2014; @geman_deep_2018]. ML methods are effective at recognizing and highlighting patterns in complex microbial datasets. However, currently the field's use of ML lacks clarity and consistency on which methods are used and how these methods are implemented. Moreover, there is a lack of deliberation on why a particular ML model is utilized. Recently, there is a trend towards using more complex ML models such as random forest, extreme gradient boosting and neural networks without a discussion on if and how much model interpretibility is necessary for the study [@galkin_human_2018; @reiman_using_2017; @fioravanti_phylogenetic_2017; @ @geman_deep_2018]. The lack of transparency on methodology and modeling decisions can have negative consequences on model reproducibility and reliability which we need to avoid by (1) implementing consistent, accessible and transparent machine learning practices; (2) selecting ML models that reflect the goal of the study as it will inform the expectations of model accuracy, complexity, interpretibility and computational efficiency.

To showcase transparent ML methodologies and to shed light on how much ML model selection can affect modeling results, we performed an empirical analysis comparing several different ML models using the same dataset. We used a previously published colorectal cancer (CRC) study [@baxter_dna_2016] which had fecal 16S rRNA gene sequences and human hemoglobin concentrations from `r paste(nrow(shared))` patients. We built ML models using fecal 16S rRNA gene sequences and human hemoglobin concentrations to predict patients with normal or screen relevant neoplasias (SRN) disease status. The study had `r paste(sum(sum(meta$dx=="normal")+sum(meta$Dx_Bin=="Adenoma")))` normal and `r paste(sum(sum(meta$dx=="cancer")+sum(meta$Dx_Bin=="adv Adenoma")))` SRN samples. We established modeling pipelines for L2-regularized logistic regression, L1 and L2 support vector machines (SVM) with linear and radial basis function kernels, a decision tree, random forest and XGBoost which increase in complexity and decrease in interpretability respectively. Our ML pipeline performed 100 data-splits and utilized held-out test data to evaluate generalization and predicton performance of each ML model. The mean AUROC varied from `r paste(round(test_all[performance_index[1],3], 2))` (std ± `r paste(round(test_all[performance_index[1],4], 2))`) to `r paste(round(test_all[performance_index[7],3], 2))` (std ± `r paste(round(test_all[performance_index[7],4], 2))`). Random Forest and XGBoost had the highest mean AUROC for detecting SRN. Despite the simplicity, the L1-regularized linear kernel SVM followed Random Forest and XGBoost in performance. In terms of computational efficiency, L2 SVM with linear kernel trained the fastest (`r paste(round(walltime_df[walltime_index[1],2], 3))` hours, std ± `r paste(round(walltime_df[walltime_index[1],3], 3))`), while XGBoost took the longest (`r paste(round(walltime_df[walltime_index[6],2], 3))` hours, std ± `r paste(round(walltime_df[walltime_index[6],3], 3))`). We found that mean cross-validation and testing AUROC could vary by as much as `r paste(round(difference[difference_index[7]], 3))`, which highlights the importance of a seperate held-out test set for evaluation. Aside from evaluating generalization and classification performances for each of these models, this study established standards for modeling pipelines of microbiome-associated machine learning models.

@@

We established modeling pipelines for binary classification with L2-regularized logistic regression, L1 and L2 support vector machines (SVM) with linear and radial basis function kernels, a decision tree, random forest and XGBoost to emphasize the differences in  model accuracy, complexity, interpretibility, computational efficiency and scalability due to  model selection. We randomly split the data into stratified training and test sets 100 times with a 80/20 split [Figure 1]. The training set consisted of 393 patients (209 SRN), while the test set was composed of 97 patients (52 SRN). The training data was used for training purposes and validation of parameter selection, and the test set was used for evaluation purposes. Validation of parameter selection was repeated 100 times to get a robust parameter selection for each model [Figure 1].

@@

We evaluated the prediction performance of seven binary classification models when applied to held-out test data over 100 data-splits using AUROC as classification performance metric. Random Forest and XGBoost had the highest mean AUROC for detecting SRN, `r paste(round(mean(rf$test_aucs), 3))` (std ± `r paste(round(sd(rf$test_aucs), 3))`) and `r paste(round(mean(xgboost$test_aucs), 3))` (std ± `r paste(round(sd(xgboost$test_aucs), 3))`) respectively [Figure 2]. L1 linear SVM and decision tree had significantly lower AUROC values, `r paste(round(mean(l1svm$test_aucs), 3))` (std ± `r paste(round(sd(l1svm$test_aucs), 3))`) and `r paste(round(mean(dt$test_aucs), 3))` (std ± `r paste(round(sd(dt$test_aucs), 3))`)  [Figure 2]. However, they had significanly higher performances than L2 linear SVM, RBF SVM and L2 logistic regression which had mean AUROC values of `r paste(round(mean(l2svm$test_aucs), 3))` (std ± `r paste(round(sd(l2svm$test_aucs), 3))`),  `r paste(round(mean(rbf$test_aucs), 3))` (std ± `r paste(round(sd(rbf$test_aucs), 3))`) and `r paste(round(mean(logit$test_aucs), 3))` (std ± `r paste(round(sd(logit$test_aucs), 3))`) respectively [Figure 2]. We also evaluated the generalization performance of each classifier by comparing their mean cross-validation AUROC and mean testing AUROC. We found a statistically significant difference for classifiers L2 support vector machines (SVM) with linear and radial basis function kernels and L2 logistic regression (p-values = `r paste(format(l2svm_diff$p.value, digits=2))`, `r paste(format(rbf_diff$p.value, digits=2))` and `r paste(format(logit_diff$p.value, digits=2))` respectively). These differences were `r paste(round(difference_model[3,1], 3))`, `r paste(round(difference_model[6,1], 3))` and `r paste(round(difference_model[4,1], 3))`, respectively [Figure 2] .

@@

We interpreted the feature weights of L1 and L2 SVM with linear kernel and regression coefficients of L2 logistic regression using the training data. To get a sense of the importance of each feature we randomly subsampled 80% of the data 100 times and trained 100 models; this resulted in 100 different sets of feature weights. We reported the mean weights of the 10 most important features [Figure 4].

@@

Linear models trained faster than non-linear models. L2 logistic Rregression and L1 and L2 SVM with linear kernel had training times of `r paste(60*(round(walltime_df[walltime_index[1],2], 2)))` min, (std ± `r paste(60*(round(walltime_df[walltime_index[1],3], 2)))`), `r paste(60*(round(walltime_df[walltime_index[2],2], 2)))` min, (std ± `r paste(60*(round(walltime_df[walltime_index[2],3], 2)))`) and `r paste(60*(round(walltime_df[walltime_index[3],2], 2)))` min, (std ± `r paste(60*(round(walltime_df[walltime_index[3],3], 2)))`) respectively. Whereas, SVM with radial basis function kernel, decision tree, random forest and xgboost had training times of `r paste(round(walltime_df[walltime_index[4],2], 2))` hrs, (std ± `r paste(round(walltime_df[walltime_index[4],3], 2))`), `r paste((round(walltime_df[walltime_index[5],2], 2)))` hrs, (std ± `r paste((round(walltime_df[walltime_index[5],3], 2)))`), `r paste(round(walltime_df[walltime_index[6],2], 2))` hrs, (std ± `r paste(round(walltime_df[walltime_index[6],3], 2))`) and `r paste(round(walltime_df[walltime_index[7],2], 2))` hrs, (std ± `r paste(round(walltime_df[walltime_index[7],3], 2))`), respectively.

>>>>

# Merge metadata and OTU table.
# Group advanced adenomas and cancers together as cancer and normal, high risk normal and non-advanced adenomas as normal
# Then remove the sample ID column
data <- inner_join(meta, shared, by=c("sample"="Group")) %>%
  mutate(dx = case_when(
    Dx_Bin== "Adenoma" ~ "normal",
    Dx_Bin== "Normal" ~ "normal",
    Dx_Bin== "High Risk Normal" ~ "normal",
    Dx_Bin== "adv Adenoma" ~ "cancer",
    Dx_Bin== "Cancer" ~ "cancer"
  )) %>%
  select(-sample, -Dx_Bin) %>%
  drop_na()
# We want the diagnosis column to be a factor
data$dx <- factor(data$dx)

@@

# ------- 1. In a loop read the files as delim
# --------2. get the model name for each with get_model_name() 
#---------3. Add a column with model name to each delim
result <- list()

for(file in walltime_files){
  model_walltime <- read_files(file) %>%  
  #summarise_walltime() %>% 
  mutate(model=get_model_name(file))
  result[[length(result)+1]] <- model_walltime
}

# ------- 1. If models are L1 or L2 Linear SVM or L2 Logit then the walltime is in minutes
# --------2. If models are Random Forest or XGBoost then the walltime is in days
#---------3. If other models, the walltime is in hours.
#---------4. This loop converts all to hours.
min_fixed_result <- list()
for(i in result){
  i$x <- i$x/3600 # The walltimes were saved as seconds we covert to hours
  min_fixed_result[[length(min_fixed_result)+1]] <- i
}
# Bind all delims with model name and walltime together
# Summarise mean walltime
walltime_df <- bind_rows(min_fixed_result) %>% 
  group_by(model) %>% 
  summarise(mean_walltime=mean(x), sd_walltime=sd(x))

@@

Advances in sequencing technology and decreasing costs of generating 16S rRNA gene sequences have allowed rapid exploration of human associated microbiome and its health implications. Currently, the human microbiome field is growing at an unprecedented rate and as a result, there is an increasing demand for methods that identify associations between members of the microbiome and human health. However, this is an undertaking as human associated microbial communities are remarkably complex and uneven. It is unlikely that a single species can explain a disease. Instead, subsets of those communities, in relation to one another and to their host, account for the differences in health outcomes. Machine learning (ML) methods are effective at recognizing and highlighting patterns in complex microbial datasets. Therefore, researchers have started to explore the utility of ML models that use microbiota associated biomarkers to predict human health and to understand the microbial ecology of diseases such as liver cirrhosis, colorectal cancer, inflammatory bowel diseases (IBD), obesity, and type 2 diabetes [@zeller_potential_2014; @zackular_human_2014; @baxter_dna_2016; @baxter_microbiota-based_2016; @hale_shifts_2017; @pasolli_machine_2016; @sze_looking_2016; @walters_meta-analyses_2014; @vazquez-baeza_guiding_2018; @qin_alterations_2014; @geman_deep_2018].  However, currently the field's use of ML lacks clarity and consistency on which methods are used and how these methods are implemented. Moreover, there is a lack of deliberation on why a particular ML model is utilized. Recently, there is a trend towards using more complex ML models such as random forest, extreme gradient boosting and neural networks without a discussion on if and how much model interpretibility is necessary for the study [@galkin_human_2018; @reiman_using_2017; @fioravanti_phylogenetic_2017; @ @geman_deep_2018]. The lack of transparency on modeling methodology and model selection negatively impact model reproducibility and reliability. We need to strive toward better machine learning practices by (1) implementing consistent, accessible and transparent machine learning methods; (2) selecting ML models that reflect the goal of the study as it will inform our expectations of model accuracy, complexity, interpretibility and computational efficiency.

To showcase transparent ML methodologies and to shed light on how much ML model selection can affect modeling results, we performed an empirical analysis comparing several different ML models using the same dataset. We used a previously published colorectal cancer (CRC) study [@baxter_dna_2016] which had fecal 16S rRNA gene sequences and human hemoglobin concentrations from `r paste(nrow(shared))` patients. We built ML models using fecal 16S rRNA gene sequences and human hemoglobin concentrations to predict patients with normal colons or patients with colonic tumors which are called screen relevant neoplasias (SRN). The study had `r paste(sum(sum(meta$dx=="normal")+sum(meta$Dx_Bin=="Adenoma")))` normal and `r paste(sum(sum(meta$dx=="cancer")+sum(meta$Dx_Bin=="adv Adenoma")))` SRN samples. We established modeling pipelines for L2-regularized logistic regression, L1 and L2 support vector machines (SVM) with linear and radial basis function kernels, a decision tree, random forest and XGBoost which are ML models that increase in complexity respectively. Our ML pipeline utilized held-out test data to evaluate generalization and prediction performance of each ML model. The mean test AUROC varied from `r paste(round(test_all[performance_index[1],3], 2))` (std ± `r paste(round(test_all[performance_index[1],4], 2))`) to `r paste(round(test_all[performance_index[7],3], 2))` (std ± `r paste(round(test_all[performance_index[7],4], 2))`). Random Forest and XGBoost had the highest mean AUROC for detecting SRN. Despite the simplicity, the L1-regularized linear kernel SVM followed Random Forest and XGBoost in performance. In terms of computational efficiency, L2 logistic regression trained the fastest (`r paste(round(walltime_df[walltime_index[1],2], 3))` hours, std ± `r paste(round(walltime_df[walltime_index[1],3], 3))`), while XGBoost took the longest (`r paste(round(walltime_df[walltime_index[7],2], 3))` hours, std ± `r paste(round(walltime_df[walltime_index[6],3], 3))`). We found that mean cross-validation and testing AUROC varied only `r paste(round(difference[difference_index[7]], 3))`, which highlights the importance of a seperate held-out test set and consistent preprocessing/splitting data for evaluation. Aside from evaluating generalization and classification performances for each of these models, this study established standards for modeling pipelines of microbiome-associated machine learning models.

@@

The data has `r paste(ncol(data)-1)` features (`r paste(ncol(data)-2)` OTUs and FIT) and a two-class label that defines the colonic health of the patients (SRN or normal). We established modeling pipelines for binary classification with L2-regularized logistic regression, L1 and L2 support vector machines (SVM) with linear and radial basis function kernels, a decision tree, random forest and XGBoost to emphasize the differences in  model accuracy, complexity, interpretibility and computational efficiency due to  model selection. We randomly split the data into stratified training/validation and test sets 100 times with a 80/20 proportion [Figure 1]. For each data-split, training/validation set consisted of 393 patients (209 SRN), while the test set was composed of 97 patients (52 SRN). The training/validation data was used for training purposes and validation of hyperparameter selection, and the test set was used for evaluation purposes. Validation of hyperparameter selection was performed by randomly splitting the training/validation set into stratified training and validation sets 100 times with a 80/20 proportion [Figure 1]. For L2-regularized logistic regression, L1 and L2 support vector machines (SVM) with linear and radial basis function kernels we tuned the __cost__ hyperparameter which determines the regularization strength where smaller values specify stronger regularization. For SVM with radial basis function kernel we also tuned __sigma__ hyperparameter which determines the reach of a single training instance where for a high value of sigma, the SVM decision boundary will be dependent on the points that are closest to the decision boundary. For the decision tree model, we tuned the __depth of the tree__ where deeper the tree, the more splits it has. For random forest, we tuned the __number of features__ to consider when looking for the best tree split. For xgboost, we tuned for __learning rate__ and the __fraction of samples__ to be used for fitting the individual base learners. We validated the prediction performances of each hyperparameter setting (Figure S1].

@@

We evaluated the prediction performance of seven binary classification models when applied to held-out test data over 100 data-splits using AUROC as classification performance metric. Random Forest and XGBoost had the highest mean AUROC for detecting SRN, `r paste(round(mean(rf$test_aucs), 3))` (std ± `r paste(round(sd(rf$test_aucs), 3))`) and `r paste(round(mean(xgboost$test_aucs), 3))` (std ± `r paste(round(sd(xgboost$test_aucs), 3))`) respectively [Figure 2]. L1 linear SVM had significantly lower AUROC values, `r paste(round(mean(l1svm$test_aucs), 3))` (std ± `r paste(round(sd(l1svm$test_aucs), 3))`) and `r paste(round(mean(dt$test_aucs), 3))` (std ± `r paste(round(sd(dt$test_aucs), 3))`)  [Figure 2]. However, it also had significanly higher performances than L2 linear SVM, RBF SVM, L2 logistic regression and decision tree which had mean AUROC values of `r paste(round(mean(l2svm$test_aucs), 3))` (std ± `r paste(round(sd(l2svm$test_aucs), 3))`),  `r paste(round(mean(rbf$test_aucs), 3))` (std ± `r paste(round(sd(rbf$test_aucs), 3))`), `r paste(round(mean(logit$test_aucs), 3))` (std ± `r paste(round(sd(logit$test_aucs), 3))`) and `r paste(round(mean(dt$test_aucs), 3))` (std ± `r paste(round(sd(dt$test_aucs), 3))`), respectively [Figure 2]. We also evaluated the generalization performance of each classifier by comparing their mean cross-validation AUROC and mean testing AUROC. We found that mean cross-validation and testing AUROC difference for each model was below 0.01.

@@

We interpreted the feature weights of L1 and L2 SVM with linear kernel and regression coefficients of L2 logistic regression using the training data. To get a sense of the most important features in linear models, we calculated the mean weights of all the features over the 100 data-splits for each model. [Figure 4]. In the three linear models, FIT was the feature that drove the detection of SRNs, followed by eight OTUs which belong to family *Lachnospiraceae* and *Ruminococcaceae* (four with positive signs; Otu01239, Otu00742, Otu00008 Otu00659, and four with negative signs; Otu00015, Otu000150, Otu00609, Otu00629). To get the importance of features in a decision tree, we used gini importance which allows us to use the improvement in the split-criterion is the importance measure attributed to the splitting variable

@@

Linear models trained faster than non-linear models. L2 logistic Rregression and L1 and L2 SVM with linear kernel had training times of `r paste(60*(round(walltime_df[4,2], 2)))` min, (std ± `r paste(60*(round(walltime_df[4,3], 2)))`), `r paste(60*(round(walltime_df[2,2], 2)))` min, (std ± `r paste(60*(round(walltime_df[2,3], 2)))`) and `r paste(60*(round(walltime_df[3,2], 2)))` min, (std ± `r paste(60*(round(walltime_df[3,3], 2)))`) respectively. Whereas, SVM with radial basis function kernel, decision tree, random forest and xgboost had training times of `r paste(round(walltime_df[6,2], 2))` hrs, (std ± `r paste(round(walltime_df[6,3], 2))`), `r paste((round(walltime_df[1,2], 2)))` hrs, (std ± `r paste((round(walltime_df[1,3], 2)))`), `r paste(round(walltime_df[5,2], 2))` hrs, (std ± `r paste(round(walltime_df[5,3], 2))`) and `r paste(round(walltime_df[7,2], 2))` hrs, (std ± `r paste(round(walltime_df[7,3], 2))`), respectively [Figure 4].

### 2019-03-13 submission/manuscript.tex
update

Advances in sequencing technology and decreasing costs of generating 16S
rRNA gene sequences have allowed rapid exploration of human associated
microbiome and its health implications. Currently, the human microbiome
field is growing at an unprecedented rate and as a result, there is an
increasing demand for methods that identify associations between members
of the microbiome and human health. However, this is an undertaking as
human associated microbial communities are remarkably complex and
uneven. It is unlikely that a single species can explain a disease.
Instead, subsets of those communities, in relation to one another and to
their host, account for the differences in the health outcomes.
Therefore, researchers have started to explore the utility of machine
learning (ML) models that use microbiota associated biomarkers to
predict human health and to understand the microbial ecology of diseases
such as liver cirrhosis, colorectal cancer, inflammatory bowel diseases
(IBD), obesity, and type 2 diabetes (1--11). ML methods are effective at
recognizing and highlighting patterns in complex microbial datasets.
However, currently the field's use of ML lacks clarity and consistency
on which methods are used and how these methods are implemented.
Moreover, there is a lack of deliberation on why a particular ML model
is utilized. Recently, there is a trend towards using more complex ML
models such as random forest, extreme gradient boosting and neural
networks without a discussion on if and how much model interpretibility
is necessary for the study (11--14). The lack of transparency on
methodology and modeling decisions can have negative consequences on
model reproducibility and reliability which we need avoid by (1)
implementing consistent, accessible and transparent machine learning
practices; (2) selecting ML models that reflect the goal of the study as
it will inform the expectations of model accuracy, complexity,
interpretibility and computational efficiency.

To showcase transparent ML methodologies and to shed light on how much
ML model selection can affect modeling results, we performed an
empirical analysis comparing several different ML models using the same
dataset. We used a previously published colorectal cancer (CRC) study
(3) which had fecal 16S rRNA gene sequences and human hemoglobin
concentrations from 490 patients. We built ML models using fecal 16S
rRNA gene sequences and human hemoglobin concentrations to predict
patients with normal or screen relevant neoplasias (SRN) disease status.
The study had 261 normal and 229 SRN samples. We established modeling
pipelines for L2-regularized logistic regression, L1 and L2 support
vector machines (SVM) with linear and radial basis function kernels, a
decision tree, random forest and XGBoost which increase in complexity
and decrease in interpretability respectively. Our ML pipeline performed
100 data-splits and utilized held-out test data to evaluate
generalization and predicton performance of each ML model. The mean
AUROC varied from 0.68 (std ± 0.05) to 0.82 (std ± 0.04). Random Forest
and XGBoost had the highest mean AUROC for detecting SRN. Despite the
simplicity, the L1-regularized linear kernel SVM followed Random Forest
and XGBoost in performance. In terms of computational efficiency, L2 SVM
with linear kernel trained the fastest (0.127 hours, std ± 0.01), while
XGBoost took the longest (82.47 hours, std ± 17.045). We found that mean
cross-validation and testing AUROC could vary by as much as 0.061, which
highlights the importance of a seperate held-out test set for
evaluation. Aside from evaluating generalization and classification
performances for each of these models, this study established standards
for modeling pipelines of microbiome-associated machine learning models.

@@

We established modeling pipelines for binary classification with
L2-regularized logistic regression, L1 and L2 support vector machines
(SVM) with linear and radial basis function kernels, a decision tree,
random forest and XGBoost to emphasize the differences in model
accuracy, complexity, interpretibility, computational efficiency and
scalability due to model selection. We randomly split the data into
stratified training and test sets 100 times with a 80/20 split {[}Figure
1{]}. The training set consisted of 393 patients (209 SRN), while the
test set was composed of 97 patients (52 SRN). The training data was
used for training purposes and validation of parameter selection, and
the test set was used for evaluation purposes. Validation of parameter
selection was repeated 100 times to get a robust parameter selection for
each model {[}Figure 1{]}.

@@

We evaluated the prediction performance of seven binary classification
models when applied to held-out test data over 100 data-splits using
AUROC as classification performance metric. Random Forest and XGBoost
had the highest mean AUROC for detecting SRN, 0.816 (std ± 0.039) and
0.814 (std ± 0.04) respectively {[}Figure 2{]}. L1 linear SVM and
decision tree had significantly lower AUROC values, 0.75 (std ± 0.045)
and 0.741 (std ± 0.038) {[}Figure 2{]}. However, they had significanly
higher performances than L2 linear SVM, RBF SVM and L2 logistic
regression which had mean AUROC values of 0.68 (std ± 0.052), 0.684 (std
± 0.049) and 0.677 (std ± 0.053) respectively {[}Figure 2{]}. We also
evaluated the generalization performance of each classifier by comparing
their mean cross-validation AUROC and mean testing AUROC. We found a
statistically significant difference for classifiers L2 support vector
machines (SVM) with linear and radial basis function kernels and L2
logistic regression (p-values = 2.4e-16, 3.5e-13 and 9.4e-20
respectively). These differences were 0.052, 0.047 and 0.061,
respectively {[}Figure 2{]} .

@@

@@

Linear models trained faster than non-linear models. L2 logistic
Rregression and L1 and L2 SVM with linear kernel had training times of
7.8 min, (std ± 0.6), 9.6 min, (std ± 0.6) and 13.2 min, (std ± 1.2)
respectively. Whereas, SVM with radial basis function kernel, decision
tree, random forest and xgboost had training times of 2.21 hrs, (std ±
0.25), 4.2 hrs, (std ± 0.49), 82.47 hrs, (std ± 17.05) and 116.83 hrs,
(std ± 17.24), respectively.

>>>>

Advances in sequencing technology and decreasing costs of generating 16S
rRNA gene sequences have allowed rapid exploration of human associated
microbiome and its health implications. Currently, the human microbiome
field is growing at an unprecedented rate and as a result, there is an
increasing demand for methods that identify associations between members
of the microbiome and human health. However, this is an undertaking as
human associated microbial communities are remarkably complex and
uneven. It is unlikely that a single species can explain a disease.
Instead, subsets of those communities, in relation to one another and to
their host, account for the differences in health outcomes. Machine
learning (ML) methods are effective at recognizing and highlighting
patterns in complex microbial datasets. Therefore, researchers have
started to explore the utility of ML models that use microbiota
associated biomarkers to predict human health and to understand the
microbial ecology of diseases such as liver cirrhosis, colorectal
cancer, inflammatory bowel diseases (IBD), obesity, and type 2 diabetes
(1--11). However, currently the field's use of ML lacks clarity and
consistency on which methods are used and how these methods are
implemented. Moreover, there is a lack of deliberation on why a
particular ML model is utilized. Recently, there is a trend towards
using more complex ML models such as random forest, extreme gradient
boosting and neural networks without a discussion on if and how much
model interpretibility is necessary for the study (11--14). The lack of
transparency on modeling methodology and model selection negatively
impact model reproducibility and reliability. We need to strive toward
better machine learning practices by (1) implementing consistent,
accessible and transparent machine learning methods; (2) selecting ML
models that reflect the goal of the study as it will inform our
expectations of model accuracy, complexity, interpretibility and
computational efficiency.

To showcase transparent ML methodologies and to shed light on how much
ML model selection can affect modeling results, we performed an
empirical analysis comparing several different ML models using the same
dataset. We used a previously published colorectal cancer (CRC) study
(3) which had fecal 16S rRNA gene sequences and human hemoglobin
concentrations from 490 patients. We built ML models using fecal 16S
rRNA gene sequences and human hemoglobin concentrations to predict
patients with normal colons or patients with colonic tumors which are
called screen relevant neoplasias (SRN). The study had 261 normal and
229 SRN samples. We established modeling pipelines for L2-regularized
logistic regression, L1 and L2 support vector machines (SVM) with linear
and radial basis function kernels, a decision tree, random forest and
XGBoost which are ML models that increase in complexity respectively.
Our ML pipeline utilized held-out test data to evaluate generalization
and prediction performance of each ML model. The mean test AUROC varied
from 0.73 (std ± 0.04) to 0.82 (std ± 0.03). Random Forest and XGBoost
had the highest mean AUROC for detecting SRN. Despite the simplicity,
the L1-regularized linear kernel SVM followed Random Forest and XGBoost
in performance. In terms of computational efficiency, L2 logistic
regression trained the fastest (0.001 hours, std ± 0.001), while XGBoost
took the longest (0.231 hours, std ± 0). We found that mean
cross-validation and testing AUROC varied only 0.007, which highlights
the importance of a seperate held-out test set and consistent
preprocessing/splitting data for evaluation. Aside from evaluating
generalization and classification performances for each of these models,
this study established standards for modeling pipelines of
microbiome-associated machine learning models.

@@

The data has 6921 features (6920 OTUs and FIT) and a two-class label
that defines the colonic health of the patients (SRN or normal). We
established modeling pipelines for binary classification with
L2-regularized logistic regression, L1 and L2 support vector machines
(SVM) with linear and radial basis function kernels, a decision tree,
random forest and XGBoost to emphasize the differences in model
accuracy, complexity, interpretibility and computational efficiency due
to model selection. We randomly split the data into stratified
training/validation and test sets 100 times with a 80/20 proportion
{[}Figure 1{]}. For each data-split, training/validation set consisted
of 393 patients (209 SRN), while the test set was composed of 97
patients (52 SRN). The training/validation data was used for training
purposes and validation of hyperparameter selection, and the test set
was used for evaluation purposes. Validation of hyperparameter selection
was performed by randomly splitting the training/validation set into
stratified training and validation sets 100 times with a 80/20
proportion {[}Figure 1{]}. For L2-regularized logistic regression, L1
and L2 support vector machines (SVM) with linear and radial basis
function kernels we tuned the \textbf{cost} hyperparameter which
determines the regularization strength where smaller values specify
stronger regularization. For SVM with radial basis function kernel we
also tuned \textbf{sigma} hyperparameter which determines the reach of a
single training instance where for a high value of sigma, the SVM
decision boundary will be dependent on the points that are closest to
the decision boundary. For the decision tree model, we tuned the
\textbf{depth of the tree} where deeper the tree, the more splits it
has. For random forest, we tuned the \textbf{number of features} to
consider when looking for the best tree split. For xgboost, we tuned for
\textbf{learning rate} and the \textbf{fraction of samples} to be used
for fitting the individual base learners. We validated the prediction
performances of each hyperparameter setting (Figure S1{]}.

@@

We evaluated the prediction performance of seven binary classification
models when applied to held-out test data over 100 data-splits using
AUROC as classification performance metric. Random Forest and XGBoost
had the highest mean AUROC for detecting SRN, 0.82 (std ± 0.033) and
0.814 (std ± 0.04) respectively {[}Figure 2{]}. L1 linear SVM had
significantly lower AUROC values, 0.76 (std ± 0.054) and 0.742 (std ±
0.04) {[}Figure 2{]}. However, it also had significanly higher
performances than L2 linear SVM, RBF SVM, L2 logistic regression and
decision tree which had mean AUROC values of 0.74 (std ± 0.045), 0.73
(std ± 0.041), 0.745 (std ± 0.039) and 0.742 (std ± 0.04), respectively
{[}Figure 2{]}. We also evaluated the generalization performance of each
classifier by comparing their mean cross-validation AUROC and mean
testing AUROC. We found that mean cross-validation and testing AUROC
difference for each model was below 0.01.

@@

We interpreted the feature weights of L1 and L2 SVM with linear kernel
and regression coefficients of L2 logistic regression using the training
data. To get a sense of the most important features in linear models, we
calculated the mean weights of all the features over the 100 data-splits
for each model. {[}Figure 4{]}. In the three linear models, FIT was the
feature that drove the detection of SRNs, followed by eight OTUs which
belong to family \emph{Lachnospiraceae} and \emph{Ruminococcaceae} (four
with positive signs; Otu01239, Otu00742, Otu00008 Otu00659, and four
with negative signs; Otu00015, Otu000150, Otu00609, Otu00629). To get
the importance of features in non-linear models, we

@@

Linear models trained faster than non-linear models. L2 logistic
Rregression and L1 and L2 SVM with linear kernel had training times of 0
min, (std ± 0), 13.8 min, (std ± 1.2) and 0 min, (std ± 0) respectively.
Whereas, SVM with radial basis function kernel, decision tree, random
forest and xgboost had training times of 0 hrs, (std ± 0), 0 hrs, (std ±
0), 0 hrs, (std ± 0) and 0 hrs, (std ± 0), respectively {[}Figure 4{]}.
